Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2016-01-23 cordillera-cycle.tex
Rev. Marshall: regional proxy records.

      Our palaeo-climate forcing therefore includes spatial temperature and
      precipitation grids derived from a~present-day atmospheric reanalysis
      \citep{Mesinger.etal.2006} that was previously tested against
      observational data and shown to best reproduce the steep precipitation
      gradients previously identified as necessary to model the LGM extent
      of the Cordilleran ice sheet in agreement with its geological imprint
      among four atmospheric reanalyses available over the study area
      \citep{Seguinot.etal.2014}. To mimic climate evolution through the
      last glacial cycle, these grids are simply supplemented by lapse-rate
      corrections and temperature offset time series. The latter are
      obtained by scaling six different palaeo-temperature reconstructions
      from proxy records around the globe, including two oxygen isotope
      records from Greenland ice cores \citep{Dansgaard.etal.1993,
      Andersen.etal.2004}, two oxygen isotope records from Antarctic ice
      cores \citep{Petit.etal.1999,Jouzel.etal.2007}, and two alkenone
      unsaturation index records from Northwest Pacific ocean sediment cores
      \citep{Herbert.etal.2001}. After testing the model sensitivity to
      these climate forcing time-series (Sect.~\ref{sec:results}) and to
      some of the most influential ice flow parameters
      (Sect.~\ref{sec:sens}), we then proceed to compare the model output
      to geological evidence and discuss the timing and extent of glaciation
      and the patterns of deglaciation, based on which we use the
      applicability of different records to modelling the history of the
      Cordilleran ice sheet (Sect.~\ref{sec:discussion}).

@@

      Each simulation starts from assumed ice-free conditions at
      120\,000\,\unit{years} ago (120\,\unit{ka}), and runs to the present.
      Our modelling domain of 1\,500 by 3\,000\,\unit{km} encompasses
      the entire area covered by the Cordilleran ice sheet at the LGM
      (Fig.~\ref{fig:locmap}). The simulations were run on two distinct
      grids, using a~lower horizontal resolution of 10\,\unit{km}, and
      a~higher horizontal resolution of 5\,\unit{km}. These computations
      were performed on 16 to 128 computing cores at the Swedish National
      Supercomputing Centre.

@@

      Numerical simulations of the Cordilleran ice sheet through the last
      glacial cycle presented in this study consistently produce two glacial
      maxima during MIS~4 (62.2--56.9\,\unit{ka},
      3.5--8.7\,\unit{m}\,s.l.e.) and MIS~2 (23.1--16.8\,\unit{ka},
      8.4--9.1\,\unit{m}\,s.l.e.), two periods corresponding to documented
      extensive glaciations. This result is independent of the
      palaeo-temperature record used among the six selected for this study,
      and thus can be regarded as a~robust model output, which broadly
      matches geological evidence. However, the timing of the two glaciation
      peaks depends sensitively on which climate record is used to drive the
      model. The timing of the LGM is best reproduced by the EPICA and
      Vostok Antarctic ice core records. It occurs about 2\,\unit{ka} too
      early in the simulation forced by the GRIP ice core record, and occurs
      even earlier in all other simulations. The mismatch is largest for the
      two Northwest Pacific ODP palaeo-temperature records, which are
      affected by the weakening of the California current during the LGM.

@@

\bibitem[{M\"{u}ller et~al.(1998)M\"{u}ller, Kirst, Ruhland, von Storch, and Rosell-Mel{\'{e}}}]{Muller.etal.1998}
M\"{u}ller, P.~J., Kirst, G., Ruhland, G., von Storch, I., and Rosell-Mel{\'{e}}, A.: Calibration of the alkenone paleotemperature index $U^{K'}_{37}$ based on core-tops from the eastern South Atlantic and the global ocean (60$^\circ$N-60$^\circ$S), Geochim. Cosmochim. Ac., 62, 1757--1772,
doi:\href{http://dx.doi.org/10.1016/s0016-7037(98)00097-0}{10.1016/s0016-7037(98)00097-0}, 1998.

@@

\bibitem[{Prahl and Wakeham(1987)}]{Prahl.Wakeham.1987}
Prahl, F.~G. and Wakeham, S.~G.: Calibration of unsaturation patterns in long-chain ketone compositions for palaeotemperature assessment, Nature, 330, 367--369,
doi:\href{http://dx.doi.org/10.1038/330367a0}{10.1038/330367a0}, 1987.

\bibitem[{Prahl et~al.(1988)Prahl, Muehlhausen, and Zahnle}]{Prahl.etal.1988}
Prahl, F.~G., Muehlhausen, L.~A., and Zahnle, D.~L.: Further evaluation of long-chain alkenones as indicators of paleoceanographic conditions, Geochim. Cosmochim. Ac., 52, 2303--2310,
doi:\href{http://dx.doi.org/10.1016/0016-7037(88)90132-9}{10.1016/0016-7037(88)90132-9}, 1988.

@@

\bibitem[{Shackleton(1967)}]{Shackleton.1967}
Shackleton, N.: Oxygen Isotope Analyses and Pleistocene Temperatures Re-assessed, Nature, 215, 15--17,
doi:\href{http://dx.doi.org/10.1038/215015a0}{10.1038/215015a0}, 1967.

>>>>

      Our palaeo-climate forcing therefore includes spatial temperature and
      precipitation grids derived from a~present-day atmospheric reanalysis
      \citep{Mesinger.etal.2006} that was previously tested against
      observational data and shown to best reproduce the steep precipitation
      gradients previously identified as necessary to model the LGM extent
      of the Cordilleran ice sheet in agreement with its geological imprint
      among four atmospheric reanalyses available over the study area
      \citep{Seguinot.etal.2014}. To mimic climate evolution through the
      last glacial cycle, these grids are simply supplemented by lapse-rate
      corrections and temperature offset time series. The latter are
      obtained by scaling six different palaeo-temperature reconstructions
      from proxy records around the globe, including two oxygen isotope
      records from Greenland ice cores \citep{Dansgaard.etal.1993,
      Andersen.etal.2004}, two oxygen isotope records from Antarctic ice
      cores \citep{Petit.etal.1999,Jouzel.etal.2007}, and two alkenone
      unsaturation index records from Northwest Pacific ocean sediment cores
      \citep{Herbert.etal.2001},

      Although these proxy records were all obtained outside the model
      domain, more regional palaeo-temperature reconstructions spanning over
      the last glacial cycle are lacking. For instance, the Mount Logan ice
      core oxygen isotope record covers only the last 30\,000\,\unit{years}
      (30\,\unit{ka}) and has been interpreted as a proxy for source region
      rather than for palaeo-temperature \citep{Fisher.etal.2004,
      Fisher.etal.2008}. Sea-surface temperatures have been reconstructed
      offshore Vancouver Island from alkenone unsaturation indices over the
      last 16\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Kienast.McKay.2001}, and
      from the Mg/Ca ratio in planktonic foraminifera over the period from
      10 to ca.~50\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Taylor.etal.2014,
      Taylor.etal.2015}, but these records cover only parts of the last
      glacial cycle.

      After testing the model sensitivity to
      these climate forcing time-series (Sect.~\ref{sec:results}) and to
      some of the most influential ice flow parameters
      (Sect.~\ref{sec:sens}), we then proceed to compare the model output
      to geological evidence and discuss the timing and extent of glaciation
      and the patterns of deglaciation, based on which we use the
      applicability of different records to modelling the history of the
      Cordilleran ice sheet (Sect.~\ref{sec:discussion}).

@@

      Each simulation starts from assumed ice-free conditions at
      120\,\unit{ka}, and runs to the present.
      Our modelling domain of 1\,500 by 3\,000\,\unit{km} encompasses
      the entire area covered by the Cordilleran ice sheet at the LGM
      (Fig.~\ref{fig:locmap}). The simulations were run on two distinct
      grids, using a~lower horizontal resolution of 10\,\unit{km}, and
      a~higher horizontal resolution of 5\,\unit{km}. These computations
      were performed on 16 to 128 computing cores at the Swedish National
      Supercomputing Centre.

@@

      Numerical simulations of the Cordilleran ice sheet through the last
      glacial cycle presented in this study consistently produce two glacial
      maxima during MIS~4 (62.2--56.9\,\unit{ka},
      3.5--8.7\,\unit{m}\,s.l.e.) and MIS~2 (23.1--16.8\,\unit{ka},
      8.4--9.1\,\unit{m}\,s.l.e.), two periods corresponding to documented
      extensive glaciations. This result is independent of the
      palaeo-temperature record used among the six selected for this study,
      and thus can be regarded as a~robust model output, which broadly
      matches geological evidence. However, the timing of the two glaciation
      peaks depends sensitively on which climate record is used to drive the
      model. The timing of the LGM is best reproduced by the EPICA and
      Vostok Antarctic ice core records. It occurs about 2\,\unit{ka} too
      early in the simulation forced by the GRIP ice core record, and occurs
      even earlier in all other simulations. The mismatch is largest for the
      two Northwest Pacific ODP palaeo-temperature records, which are
      affected by the weakening of the California current during the LGM.
      Nevertheless, the fact that Cordilleran ice sheet dynamics are here
      modelled in best agreement to geological reconstructions using
      palaeo-temperature records from the distant Greenland and Antarctic
      ice sheets, and the significant differences remaining between our two
      preffered simulations highlight the need for more regional
      palaeo-climate reconstructions of the last glacial cycle in and around
      the North American Cordillera.

@@

@@

@@
